### 2025-03-13 paper/sections/abstract.qmd

We propose a novel training regime called Counterfactual Training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to stakeholder requirements. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for this desired end goal: Counterfactual Training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

>>>>

We propose a novel training regime termed Counterfactual Training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the stakeholder requirements. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: Counterfactual Training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

### 2025-03-13 paper/sections/conclusion.qmd

# Conclusion {#sec-conclusion}

\begin{credits}
\subsubsection{\ackname}

- Aleks
- DelftBlue
- Julia packages: Julia paper itself, PrettyTables, Flux, Plots, AlgebraOfGraphics

\subsubsection{\discintname}

The authors have no competing interests to declare that are relevant to the content of this article.
\end{credits}

>>>>

# Conclusion {#sec-conclusion}
State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans. Although post-hoc explainability approaches have attracted major research interest, these cannot guarantee that the explanations agree with the opaque model's learned representation of data. As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models. Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients. Through extensive experiments we demonstrate that counterfactual training satisfies its objectives while preserving the predictive performance of the trained models. We also find that our approach can be used to fine-tune conventionally-trained models and achieve similar gains in explainability. Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.
